Modelling Stochastic Time Delay Structures for Regression

### 2021-02-06 documentation/article/article.tex
added v1 algorithm

\begin{align}
x(t) = \left(0,0,1,0,1,0\right)
\end{align}

@@

where $X(t + T)$ is the time shifted series and $X(t)$ is the observed input. \\

In order to find the best estimate of f, we would like to find the function $f$ which maximises the joint likelihood of both Tau and e, the time-domain likelihood and the error likelihood respectively.

@@

where $\theta_{\varepsilon}$ and $\theta_{\tau}$ represent the parameters of the prediction error and time error respectively. For simplicity, assume that the tau and error distributions are independent. To begin with, the values of each individual time shift ($\tau$) are not known. In addition, the size of prediction error $\varepsilon$ can only be determined if each tau is known (because for each time shift there is a different prediction and prediction error). Therefore, our algorithm finds the best estimate of the individual time shifts in an inner optimisation, then iteratively searches for the best estimate of f in an outer optimisation step. \\

>>>>

\begin{align}
x(t) = 
\left(
\begin{array}{cccccc}
0 & 0 & 1 & 0 & 1 & 0
\end{array}  
\right)
\end{align}

@@

where $X(t + T)$ is the time shifted series and $X(t)$ is the observed input.\\

In order to find the best estimate of $f$, we would like to find the function $f$ which maximises the joint likelihood of both Tau(?) and $\varepsilon$, the time-domain likelihood and the error likelihood respectively.

@@

where $\theta_{\varepsilon}$ and $\theta_{\tau}$ represent the parameters of the prediction error and time error respectively. For simplicity, assume that the tau and error distributions are independent. To begin with, the values of each individual time shift ($\tau$) are not known. In addition, the size of prediction error $\varepsilon$ can only be determined if each tau is known (because for each time shift there is a different prediction and prediction error). Therefore, our algorithm finds the best estimate of the individual time shifts in an inner optimisation, then iteratively searches for the best estimate of $f$ in an outer optimisation step.

\section{Algorithm}

Firstly, we define a parametric form and some initial parameters to be estimated for the function $f$. For example, let's take the simple univariate linear model where the $f(x(t))$ is parameterised by $\beta$, $\sigma_{\varepsilon}$(error standard deviation)  and $\sigma_{\tau}$ (time error standard deviation). First, we initialise some starting values for each of these parameters.

Now we want to find the best possible time shift $\tau$ for each input impulse in $X(t)$. It stands to reason that the best possible time shift would be one that is not too far away from the observed impulse and also gives the best possible prediction. In this example, we can get the prediction y by simply multiplying the shifted value by its parameter $\beta$. From there we can calculate the likelihood estimate for time shift plus the prediction error.

In principle, we can then try a number of values of $\tau$ (i.e. optimise) until we maximise the likelihood for this impulse.

However, we must also consider that the impulses in $X(t)$ are not independent from each other. After shifting, it's possible that two or more effects can occur simultaneously.

This could be particularly problematic if there are multiple impulses within a short period of time, or the impulses have a distributed effect over multiple time points.

As an example consider the series

\begin{align}
x(t) = 
\left(
\begin{array}{cccccc}
0 & 0 & 1 & 1 & 0 & 0
\end{array}  
\right)
\end{align}

with $T= (1, 0)$ and $\beta = 1$. For this case,

\begin{align}
X(t + T) = 
\left(
\begin{array}{cccccc}
0 & 0 & 1 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0 & 0 
\end{array}  
\right)
\end{align}

and the effect is therefore 
\begin{align}
y = 
\left(
\begin{array}{cccccc}
0 & 0 & 0 & 2 & 0 & 0
\end{array}  
\right)
\end{align}

We must consider the effects at the same time to accurately calculate the likelihood.

As shown above, if the impulses are not independent, then we need to consider time shifts $\tau$ at the same time. Therefore, we treat the problem of finding the best time shifts as a discrete optimisation problem.

For the optimisation step we utilise two assumptions to complete the optimisation. Firstly, smaller shifts are more likely than larger shifts (proportionate to the standard deviation of the T distribution). This means that we should explore the space of smaller shifts more often than larger shifts. Secondly, impulses close to each other are more likely to be dependent on each other,  than impulses further away.

Therefore, we initialize the optimisation algorithm with all parameters set to the mean of the T distribution (e.g. zero for the zero mean gaussian). Then we randomly select a small number of impulses, the exact number could be specified manually as a hyperparameter, or inferred from the data by looking at the density of impulses in time. These impulses are randomly changed to a value drawn for the T distribution (proposal vector), and the likelihood (both time shift and prediction error) for the proposal is calculated. If the proposal likelihood is higher than the current maximum likelihood, then we update our initial estimate to the current best guess. This procedure is done in a loop, with the best guess of the set of T improving over time.

Lastly, we optimise over the set of parameters $\beta$, $\sigma_{\varepsilon}$ and $\sigma_{\tau}$, iteratively alternating between the time shift optimisation and the parameter optimisation. For the parameter optimisation, typical methods can be used such as gradient descent, genetic algorithms or annealing.  In our implementation we have used the differential evolution algorithm from the scipiy.optimize library (REF) in python, due to its ability to handle noisy objective functions. In order to improve convergence, we also standardize all input variables to the range 0-1.

Through optimisation, we find the best fit for the values of $\beta$, $\sigma_{\varepsilon}$ and $\sigma_{\tau}$. The accuracy of the final parameter estimate is relative to the ratio of the $y$-axis error and the effect size $\beta X(t)$. As $\sigma_{\varepsilon}/(\beta X(t))$ tends to infinity, the estimated value of $\sigma_{\tau}$ tends to zero. In the event that $\sigma_{\tau}=0$, the model becomes a ‘fixed lag’ model, and for the discrete gaussian case the parameter estimate $\beta$ tends to the standard linear regression coefficient.
On the other hand as $\sigma_{\varepsilon}/(\beta X(t))$ tends to zero, the estimated value of B(?) tends to the true value. Therefore, the method provides no guarantee on recovering the exact time shifts, instead only that we obtain an estimate equal to or better than standard regression.

\section{Limitations}

\section{Experiments}

\section{Future Work}

\section{Conclusion}

We have proposed a form of regression analysis for dealing with stochastic time delay problems. We have shown the feasibility of the approach and its performance on simulated data. The approach allows for better estimation and prediction when the input is affected by noise in the time domain. To our knowledge, the method is novel for this class of problem.
